Data Specialization for a Staged Lambda Calculus

### 2014-11-06 paper/splitting.tex
Just made the transition to motivating the definition of splitting in 4.2.

Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
\bbtwo\ components, these components may be interleaved in \lang\ code. The
goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
of a term $e$ in \lang\ into a pair of
terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same value 
as evaluating $e$ in \lang.

@@

Splitting a stage~\bbtwo\ term $\coltwo e A$ generates
 a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
of $e$ in place to create a residual, splitting statically isolates these computations into
$p$.  The resumer accepts the value produced by the precomputation as input
% (bound to the free variable $l$)
and executes all stage~\bbtwo\ subcomputations.   
Specifically, if $e$ in \lang\ reduces to the final value $v$, 
then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
(The resumer, bound to the output of $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})

@@

For all non-terminal stage~\bbtwo\ terms $e$ in \lang\ (except $\prev$ and $\pause$), splitting descends into $e$, recursively splitting subterms.  
The precompuation is formed by combining the resultant $n$ subterm precomputations into a tuple: $p=(p_1,\ldots,p_n)$.
The generated resumer has the same structure as $e$, with every subterm replaced by its resumer.
Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
underneath stage \bbtwo\ binders. % should probably draw a parallel to the same behavior in dynamics

@@

Note that this result carries information at {\em both} stages: 
at stage~\bbone\ it carries the value 3, and at stage~\bbtwo\ it carries the code $\letin \yhat {\mathtt{7+5}} \yhat$.

But the whole point of splitting is that its results (and their evaluated forms) are monostage.
Ergo, if we intend to compare the results of evaluation to the those of splitting, 
then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.

% Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.
To this end, we introduce two {\em masking} functions...

@@

We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
a partial value at each stage, essentially by replacing all stage \bbone\
(resp., \bbtwo) components of a partial value by \texttt{()}. The first result
of splitting must then compute the \bbone-mask of the partial value, and the
second result the \bbtwo-mask.

There is one final subtlety to splitting stage \bbone\ terms. In the example
above, there are \emph{two} flavors of stage \bbone\ subcomputations:
\texttt{1+2}, which is needed to compute the \bbone-mask, and \texttt{3+4},
a precomputation which is used by the computation of the \bbtwo-mask.

\ur{This paragraph is unintellegible to me.  What is the point?}

>>>>

Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
\bbtwo\ components, these components may be interleaved in \lang\ code. The
goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
of a term $e$ in \lang\ into a pair of
terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
Intuitively, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
as evaluating $e$ in \lang.

@@

Splitting a stage~\bbtwo\ term $\coltwo e A$ generates
 a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
of $e$ in place to create a residual, splitting statically isolates these computations into
$p$.  The resumer accepts the value produced by the precomputation as input
% (bound to the free variable $l$)
and executes all stage~\bbtwo\ subcomputations.   
Specifically, if $e$ in \lang\ reduces to the final value $v$, 
then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
(Thus, the resumer, bound to the value output by $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})

@@

For all non-terminal stage~\bbtwo\ terms $e$ in \lang\ (except $\prev$ and $\pause$), splitting descends into $e$, recursively splitting subterms.  
The precompuation is formed by combining the resultant $n$ subterm precomputations into a tuple: $p=(p_1,\ldots,p_n)$.
The generated resumer has the same structure as $e$, with every subterm replaced by its resumer.
Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
underneath stage \bbtwo\ binders.  % TODO: should probably draw a parallel to the same behavior in dynamics

@@

Note that this result carries information at {\em both} stages: 
at stage~\bbone\ it carries the value 3, and at stage~\bbtwo\ it carries the code $\letin \yhat {\mathtt{7+5}} \yhat$. 
Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms to create a resumer that reduces to the same value as the stage~\bbtwo\ carried by the partial value, and an $e1$ that not only performs the precomputation necessary for the resumer, but also generates values that match the stage~\bbone\ aspects of the partial value.

% Ergo, if we intend to compare the results of evaluation to the those of splitting, 
% then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.

% But the whole point of splitting is that its results (and their evaluated forms) are monostage.

% Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.

To concretize what it means by ``match", we introduce two {\em masking} functions, $\masko{\cdot}$ and $\maskt{\cdot}$, that project from the residual table and partial value only the parts that are relevant at a particular stage.  These functions are shown in \ref{fig:valMask}. They operate by replacing all stage \bbone\
(resp., \bbtwo) components of a partial value by \texttt{()}.

@@

%We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
%$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
%a partial value at each stage, essentially by replacing all stage \bbone\
%(resp., \bbtwo) components of a partial value by \texttt{()}.

The first result
of splitting must then compute the \bbone-mask of the partial value, and the
second result the \bbtwo-mask.

% There is one final subtlety to splitting stage \bbone\ terms. In the example
% above, there are \emph{two} flavors of stage \bbone\ subcomputations:
% \texttt{1+2}, which is needed to compute the \bbone-mask, and \texttt{3+4},
% a precomputation which is used by the computation of the \bbtwo-mask.
